Estimating duration in the presence of misclassification

### 2024-03-23 biology-data.tex
Standardise age group names

The CIS had a household-based design, meaning that households were invited to participate in the study.
From households that participated, all individuals aged 2 and over were invited to participate.
Once invited, an enrolment swab would be taken at the first visit followed by 4 further weekly visits (giving a total of 5 swabs on days 0, 7, 14, 21, 28 relative to enrolment) after which visits were monthly.
As well as a swab to test for SARS-CoV-2, participants were asked to complete a questionnaire, and some participants were asked to provide blood samples.
In reality, visits were often not on this precise schedule, and occasionally missed.
A full description of the study can be found in the study protocol~\autocite{cisProtocol}.

@@

The age groups used have varied slightly for different CIS analyses~\autocite[e.g.][]{pouwelsMRPvaccination,pouwelsCommunity,cisMethodsONS,houseInferring,walkerTracking}.
I will use the age groups: 0--10 (pre-school and primary school), 11--15 (compulsory secondary school), 16--24 (adolescents and young adults who may remain in formal education), 25--49 (younger working age adults), 50--69 (older working age adults), and 70+ (elderly).

>>>>

The CIS had a household-based design, meaning that households were invited to participate in the study.
From households that participated, all individuals aged 2 and over were invited to participate.
Once invited, an enrolment swab would be taken at the first visit followed by 4 further weekly visits (giving a total of 5 swabs on days 0, 7, 14, 21, 28 relative to enrolment) after which visits were monthly.
As well as a swab to test for SARS-CoV-2, participants were asked to complete a questionnaire, and some participants were asked to provide blood samples.
In reality, visits were often not on this precise schedule, and occasionally missed altogether.
A full description of the study can be found in the study protocol~\autocite{cisProtocol}.

@@

The age groups used have varied slightly for different CIS analyses~\autocite[e.g.][]{pouwelsMRPvaccination,pouwelsCommunity,cisMethodsONS,houseInferring,walkerTracking}.
I will use the age groups: 0--11 (pre-school and primary school), 12--16 (compulsory secondary school), 17--25 (adolescents and young adults who may remain in formal education), 26--50 (younger working age adults), 51--70 (older working age adults), and 71+ (elderly).

### 2024-03-23 transmission.tex
Standardise age group names

The incidence proportion was highest among the 11--15 and 16--25 age groups throughout the period (see \cref{transmission:fig:backcalc-ages}).
Except the 16--25 age group, all age groups show similar trends (\ie parallel lines on a log scale), meaning there was insufficient information in the data to overcome the prior which favoured this situation.
The incidence proportion in the 16--25 age group decreases faster than other age groups during November.
There are many plausible explanations for this effect.
The higher incidence in this age group prior to November means it could be an immunity effect.
Alternatively, it could be a consequence of schools remaining open in the November lockdown; 20--24 year olds rarely interact with school-aged children (see \cref{E-SEIR:fig:age-contacts})
Many other explanations are possible, and further work is required to determine causality.
\begin{figure}
    \makebox[\textwidth][c]{\includegraphics[width=.9\paperwidth]{transmission/backcalc-ages}}
    \caption[Incidence estimated using backcalculation by age group]{%
        Incidence proportion for each age group, on both natural and log scales.
    }
    \label{transmission:fig:backcalc-ages}
\end{figure}

@@

% The model will be stratified only by region and age group, so the data are aggregated to these levels.
% Only individuals aged two or older are eligible to participate in the CIS; I assume that test results in all participants aged under 11 are still representative of the whole $[0, 11)$ age group.

@@

\centering
\begin{tabular}{l|lllll}
         & $[0,16)$ & $[16,25)$ & $[25,50)$ & $[50,70)$ & $[70,\infty)$ \\
        \hline
        East Midlands & $\BetaDist(8.5, 301)$ & $\BetaDist(18, 256)$ & $\BetaDist(19, 475)$ & $\BetaDist(20, 493)$ & $\BetaDist(5, 337)$ \\
        East of England & $\BetaDist(7.9, 236)$ & $\BetaDist(21, 248)$ & $\BetaDist(35, 677)$ & $\BetaDist(26, 613)$ & $\BetaDist(5.6, 332)$ \\
        London & $\BetaDist(9.7, 131)$ & $\BetaDist(50, 328)$ & $\BetaDist(154, 1331)$ & $\BetaDist(106, 990)$ & $\BetaDist(7.6, 204)$ \\
        North East & $\BetaDist(6.1, 180)$ & $\BetaDist(12, 145)$ & $\BetaDist(14, 274)$ & $\BetaDist(14, 295)$ & $\BetaDist(4.2, 240)$ \\
        North West & $\BetaDist(12, 252)$ & $\BetaDist(38, 353)$ & $\BetaDist(72, 885)$ & $\BetaDist(44, 695)$ & $\BetaDist(6.3, 264)$ \\
        South East & $\BetaDist(8.9, 404)$ & $\BetaDist(17, 319)$ & $\BetaDist(20, 577)$ & $\BetaDist(19, 586)$ & $\BetaDist(4.4, 356)$ \\
        South West & $\BetaDist(6.9, 415)$ & $\BetaDist(11, 277)$ & $\BetaDist(12, 556)$ & $\BetaDist(11, 559)$ & $\BetaDist(4, 492)$ \\
        West Midlands & $\BetaDist(8, 167)$ & $\BetaDist(19, 151)$ & $\BetaDist(32, 475)$ & $\BetaDist(33, 504)$ & $\BetaDist(5.8, 241)$ \\
        Yorkshire and The Humber & $\BetaDist(9.2, 314)$ & $\BetaDist(19, 242)$ & $\BetaDist(27, 589)$ & $\BetaDist(18, 474)$ & $\BetaDist(4.9, 316)$ \\
    \end{tabular}
\caption{Priors for $\vec{\pi}$, the proportion of the population without prior immunity, in each age group and region combination.}
\label{SEIR:table:immunity-prior}
\end{table}
\end{landscape}

@@

The 11--16 age group (roughly corresponding to compulsory secondary school education) is estimated to have the highest incidence (see \cref{SEIR:fig:incidence}).
Incidence is smooth within each week but shows discontinuities between weeks.
This is because the parameters within $\matr{\beta}(t)$ are constant within each week.
A constant $\matr{\beta}(t)$ would lead to exponential growth following a period of transitory dynamics, although the weekly switching may mean that these transitory dynamics never end, especially when incidence is high~\autocite{rhodesConvergence}.
The estimated timing of the peak in incidence varies by region, with the most common time being late December 2020 (see \cref{SEIR:fig:peak-incidence}).
In Yorkshire and the Humber, the October peak is estimated to be greater than the winter peak; for North West there is a significant probability on either being larger.
\begin{figure}
    \includegraphics{SEIR/CIS/incidence}
    \caption[Posterior estimates of incidence]{%
        Posterior estimates (95\% CrI) of incidence by region and age.
        See main text for discussion.
    }
    \label{SEIR:fig:incidence}
\end{figure}
\begin{figure}
    \includegraphics{SEIR/CIS/p_peak}
    \caption[Posterior estimates of peak incidence timing]{%
        Posterior probability of the peak incidence for each region and age combination occurring on each day.
    }
    \label{SEIR:fig:peak-incidence}
\end{figure}

@@

The \emph{attack rate} is the proportion of the population that have been infected at least once.
In this model it is $1 - \vec{s}$, the proportion susceptible at the time the study ends.
The attack rate is highest in the $[11, 16)$ age group in the North West at 68\% (95\% CrI: 66--70\%) (see \cref{SEIR:fig:attack-rates}).
Some of these infections occurred prior to the period covered by this study; during the period of this study 23\% (95\% CrI: 21--24\%) of the population in the North West aged 11--16 was infected.
\begin{figure}
    \makebox[\textwidth][c]{\includegraphics{SEIR/CIS/attack_rates}}
    \caption{Posterior median and 95\% CrI of the attack rate in each region and age group.}
    \todo[inline]{Maybe compare this to blood donors, as used in the prior.}
    \label{SEIR:fig:attack-rates}
\end{figure}

Overall, the model fits the data well (see \cref{SEIR:fig:prev,SEIR:fig:prev-appendix}).
There are a few features of the data which are not captured by the model and the posterior estimate appears to be a compromise with the prevalence too low in one age group and too high in another.
For example, in the North East region, the 11--16 age group has its prevalence overestimated in October but the 16--25 age group has its prevalence underestimated.
\begin{figure}
    % \thisfloatpagestyle{empty}
    % \vspace{-3cm}
    \makebox[\textwidth][c]{\includegraphics[width=.9\paperwidth]{SEIR/CIS/prev_main}}
    \captionsetup{width=0.8\paperwidth}
    \captionof{figure}[SEIR prevalence goodness-of-fit]{%
        Goodness-of-fit to prevalence data by region and age for selected regions (see \cref{SEIR:fig:prev-appendix} for others).
        The central ribbon and line show the posterior median and 95\% CrI estimate for the proportion of the relevant stratum RT-PCR positive on each day.
        The lighter outer ribbon is the 95\% posterior predictive interval for the proportion of test results that are positive each week.
        The dots show the observed prevalence in CIS, aggregated by week.
        For a well-calibrated model, 95\% of the dots would be within the outer ribbon.
    }
    \label{SEIR:fig:prev}
\end{figure}

@@

The current model could be more flexible with regards to the age distribution of infections.
This would better exploit one of the strengths of the CIS: the availability of data  on all age groups.
In some regions, there appears to be a trade-off between under and overfitting in different age groups (see \cref{SEIR:fig:prev}).
For example, in the North East, most of the data points in the $[11, 16)$ age group are below the median posterior estimate, while most of the data points in the $[16, 25)$ age group are above the median posterior estimate.
This implies that the data would allow identification of a more flexible age-based model.

@@

Many of these issues are clearly demonstrated by the estimates for the North East (see \cref{transmission:fig:compare-NE}).
I previously observed the quality of the SEIR model's fit is variable between different age groups (see \cref{SEIR:fig:prev,SEIR:fig:prev-appendix}).
This is because the model is not flexible enough to capture the true age distribution of infections.
The overestimation of prevalence in the 11--16 age group and underestimation in the 16--25 age group is reflected when comparing incidence estimates to those from the backcalculation model, which have reduced bias.
\begin{figure}
    \centering \includegraphics{transmission/compare-NE}
    \caption[Comparing each models estimate of North East incidence by age.]{%
        Incidence proportion in each age group in the North East estimated by the two models discussed in this chapter.
        Posterior median and 95\% CrI shown.
        For the SEIR model, newly RT-PCR positive (entrances into the P1 compartment), rather than incidence (entrances into the E compartment) are shown for comparability with the backcalculation results.
    }
    \label{transmission:fig:compare-NE}
\end{figure}

>>>>

The incidence proportion was highest among the 12--16 and 17--25 age groups throughout the period (see \cref{transmission:fig:backcalc-ages}).
Except the 17--25 age group, all age groups show similar trends (\ie parallel lines on a log scale), meaning there was insufficient information in the data to overcome the prior which favoured this situation.
The incidence proportion in the 17--25 age group decreases faster than other age groups during November.
There are many plausible explanations for this effect.
The higher incidence in this age group prior to November means it could be an immunity effect.
Alternatively, it could be a consequence of schools remaining open in the November lockdown; 20--24 year olds rarely interact with school-aged children (see \cref{E-SEIR:fig:age-contacts})
Many other explanations are possible, and further work is required to determine causality.
\begin{figure}
    \makebox[\textwidth][c]{\includegraphics[width=.9\paperwidth]{transmission/backcalc-ages}}
    \caption[Incidence estimated using backcalculation by age group]{%
        Incidence proportion for each age group, on both natural and log scales.
    }
    \label{transmission:fig:backcalc-ages}
\end{figure}

@@

% The model will be stratified only by region and age group, so the data are aggregated to these levels.
% Only individuals aged two or older are eligible to participate in the CIS; I assume that test results in all participants aged under 11 are still representative of the whole 0--11 age group.

@@

\centering
\begin{tabular}{l|lllll}
         & 0--11 & 12--16 & 17--25 & 26--50 & 71+ \\
        \hline
        East Midlands & $\BetaDist(8.5, 301)$ & $\BetaDist(18, 256)$ & $\BetaDist(19, 475)$ & $\BetaDist(20, 493)$ & $\BetaDist(5, 337)$ \\
        East of England & $\BetaDist(7.9, 236)$ & $\BetaDist(21, 248)$ & $\BetaDist(35, 677)$ & $\BetaDist(26, 613)$ & $\BetaDist(5.6, 332)$ \\
        London & $\BetaDist(9.7, 131)$ & $\BetaDist(50, 328)$ & $\BetaDist(154, 1331)$ & $\BetaDist(106, 990)$ & $\BetaDist(7.6, 204)$ \\
        North East & $\BetaDist(6.1, 180)$ & $\BetaDist(12, 145)$ & $\BetaDist(14, 274)$ & $\BetaDist(14, 295)$ & $\BetaDist(4.2, 240)$ \\
        North West & $\BetaDist(12, 252)$ & $\BetaDist(38, 353)$ & $\BetaDist(72, 885)$ & $\BetaDist(44, 695)$ & $\BetaDist(6.3, 264)$ \\
        South East & $\BetaDist(8.9, 404)$ & $\BetaDist(17, 319)$ & $\BetaDist(20, 577)$ & $\BetaDist(19, 586)$ & $\BetaDist(4.4, 356)$ \\
        South West & $\BetaDist(6.9, 415)$ & $\BetaDist(11, 277)$ & $\BetaDist(12, 556)$ & $\BetaDist(11, 559)$ & $\BetaDist(4, 492)$ \\
        West Midlands & $\BetaDist(8, 167)$ & $\BetaDist(19, 151)$ & $\BetaDist(32, 475)$ & $\BetaDist(33, 504)$ & $\BetaDist(5.8, 241)$ \\
        Yorkshire and The Humber & $\BetaDist(9.2, 314)$ & $\BetaDist(19, 242)$ & $\BetaDist(27, 589)$ & $\BetaDist(18, 474)$ & $\BetaDist(4.9, 316)$ \\
    \end{tabular}
\caption{Priors for $\vec{\pi}$, the proportion of the population without prior immunity, in each age group and region combination.}
\label{SEIR:table:immunity-prior}
\end{table}
\end{landscape}

@@

The 12--16 age group (roughly corresponding to compulsory secondary school education) is estimated to have the highest incidence (see \cref{SEIR:fig:incidence}).
Incidence is smooth within each week but shows discontinuities between weeks.
This is because the parameters within $\matr{\beta}(t)$ are constant within each week.
A constant $\matr{\beta}(t)$ would lead to exponential growth following a period of transitory dynamics, although the weekly switching may mean that these transitory dynamics never end, especially when incidence is high~\autocite{rhodesConvergence}.
The estimated timing of the peak in incidence varies by region, with the most common time being late December 2020 (see \cref{SEIR:fig:peak-incidence}).
In Yorkshire and the Humber, the October peak is estimated to be greater than the winter peak; for North West there is a significant probability on either being larger.
\begin{figure}
    \includegraphics{SEIR/CIS/incidence}
    \caption[Posterior estimates of incidence]{%
        Posterior estimates (95\% CrI) of incidence by region and age.
        See main text for discussion.
    }
    \label{SEIR:fig:incidence}
\end{figure}
\begin{figure}
    \includegraphics{SEIR/CIS/p_peak}
    \caption[Posterior estimates of peak incidence timing]{%
        Posterior probability of the peak incidence for each region and age combination occurring on each day.
    }
    \label{SEIR:fig:peak-incidence}
\end{figure}

@@

The \emph{attack rate} is the proportion of the population that have been infected at least once.
In this model it is $1 - \vec{s}$, the proportion susceptible at the time the study ends.
The attack rate is highest in the 12--16 age group in the North West at 68\% (95\% CrI: 66--70\%) (see \cref{SEIR:fig:attack-rates}).
Some of these infections occurred prior to the period covered by this study; during the period of this study 23\% (95\% CrI: 21--24\%) of the population in the North West aged 11--16 was infected.
\begin{figure}
    \makebox[\textwidth][c]{\includegraphics{SEIR/CIS/attack_rates}}
    \caption{Posterior median and 95\% CrI of the attack rate in each region and age group.}
    \todo[inline]{Maybe compare this to blood donors, as used in the prior.}
    \label{SEIR:fig:attack-rates}
\end{figure}

Overall, the model fits the data well (see \cref{SEIR:fig:prev,SEIR:fig:prev-appendix}).
There are a few features of the data which are not captured by the model and the posterior estimate appears to be a compromise with the prevalence too low in one age group and too high in another.
For example, in the North East region, the 12--16 age group has its prevalence overestimated in October but the 17--25 age group has its prevalence underestimated.
\begin{figure}
    % \thisfloatpagestyle{empty}
    % \vspace{-3cm}
    \makebox[\textwidth][c]{\includegraphics[width=.9\paperwidth]{SEIR/CIS/prev_main}}
    \captionsetup{width=0.8\paperwidth}
    \captionof{figure}[SEIR prevalence goodness-of-fit]{%
        Goodness-of-fit to prevalence data by region and age for selected regions (see \cref{SEIR:fig:prev-appendix} for others).
        The central ribbon and line show the posterior median and 95\% CrI estimate for the proportion of the relevant stratum RT-PCR positive on each day.
        The lighter outer ribbon is the 95\% posterior predictive interval for the proportion of test results that are positive each week.
        The dots show the observed prevalence in CIS, aggregated by week.
        For a well-calibrated model, 95\% of the dots would be within the outer ribbon.
    }
    \label{SEIR:fig:prev}
\end{figure}

@@

The current model could be more flexible with regards to the age distribution of infections.
This would better exploit one of the strengths of the CIS: the availability of data  on all age groups.
In some regions, there appears to be a trade-off between under and overfitting in different age groups (see \cref{SEIR:fig:prev}).
For example, in the North East, most of the data points in the 12--16 age group are below the median posterior estimate, while most of the data points in the 17--25 age group are above the median posterior estimate.
This implies that the data would allow identification of a more flexible age-based model.

@@

Many of these issues are clearly demonstrated by the estimates for the North East (see \cref{transmission:fig:compare-NE}).
I previously observed the quality of the SEIR model's fit is variable between different age groups (see \cref{SEIR:fig:prev,SEIR:fig:prev-appendix}).
This is because the model is not flexible enough to capture the true age distribution of infections.
The overestimation of prevalence in the 12--16 age group and underestimation in the 17--25 age group is reflected when comparing incidence estimates to those from the backcalculation model, which have reduced bias.
\begin{figure}
    \centering \includegraphics{transmission/compare-NE}
    \caption[Comparing each models estimate of North East incidence by age.]{%
        Incidence proportion in each age group in the North East estimated by the two models discussed in this chapter.
        Posterior median and 95\% CrI shown.
        For the SEIR model, newly RT-PCR positive (entrances into the P1 compartment), rather than incidence (entrances into the E compartment) are shown for comparability with the backcalculation results.
    }
    \label{transmission:fig:compare-NE}
\end{figure}
